\documentclass[12pt]{report}
\usepackage{listings}
\usepackage{amsthm}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{tikz}
\usetikzlibrary{arrows}
\usepackage[T1]{fontenc}
\usepackage{float}
\usepackage{url}
\usepackage{geometry}
\usepackage{marginnote}
\usepackage{url}
\usepackage[autostyle]{csquotes}
\newtheorem*{definition}{Definition}
\theoremstyle{definition}
\newtheorem*{theorem}{Theorem}
\theoremstyle{theorem}

\lstdefinestyle{customc}{
	belowcaptionskip=1\baselineskip,
	breaklines=true,
	xleftmargin=\parindent,
    numbers=left,
    numberstyle=\ttfamily\color{gray},
	language=Java,
	showstringspaces=false,
	basicstyle=\footnotesize\ttfamily,
	keywordstyle=\bfseries\color{green!40!black},
	commentstyle=\itshape\color{purple!40!black},
	identifierstyle=\color{black},
	stringstyle=\color{orange},
}
\lstdefinestyle{customasm}{
	belowcaptionskip=1\baselineskip,
	frame=L,
	xleftmargin=\parindent,
	language=[x86masm]Assembler,
	basicstyle=\footnotesize\ttfamily,
	commentstyle=\itshape\color{purple!40!black},
}
\lstset{escapechar=@,style=customc}

\begin{document}
\begin{titlepage}
	\centering
    \null
    \vfill
    {\Large\itshape Comparing Maintainability in Software Paradigms
    \par}
    \vspace{3.0cm}
	{\scshape 
    A thesis presented \par 
    by\par
	Marc Coquand\par
	to\par
    The Department of Computer Science\par
    \vspace{0.8cm}
	for a degree of\par
    Master of Science in Engineering\par
    in the subject of\par
    Interaction Technology and Design\par}
    \vfill
    Umeå University\par
    Supervised by Anders Broberg\par
	\today\par
\end{titlepage}
\clearpage
\thispagestyle{empty}

\clearpage\newpage
\thispagestyle{empty}

\begin{abstract} 

    This study's goal is to compare approaches to functional programs and
    object-oriented programs to find how it affects maintainability and code
    quality. By looking at 3 cases, we analyze, how does a functional approach
    to software architecture compare to an OOP (Object-oriented programming)
    approach when it comes to maintainability and code quality? TO BE REPLACED
    WITH CONCLUSION

\end{abstract}

\clearpage\newpage
\thispagestyle{empty}

\section*{Acknowledgements}

I want to thank myself for my amazing work. TODO

\clearpage\newpage
\thispagestyle{empty}

\tableofcontents
\newpage

% Introduction
\chapter{Introduction: Software paradigms and complexity}

Different schools of thoughts have different approaches when it comes to
building applications. There is one that is the traditional, object oriented,
procedural way of doing it. Then there is a contender, a functional approach, as
an alternative way to build applications. When building applications testability
is of high concern to ensure that the application functions properly. By looking
at cyclomatic complexity, described in Section~\ref{cyclomaticcomplexity}, we
can find out how the different approaches affect the amount of tests we need to
write to get full branch coverage. By looking at the cognitive dimensions,
described in Section~\ref{cognitivedimensions}, we can find how the two
approaches affect the mental complexity for the developer. So by looking at
different case studies this study aims to find if the different software
paradigms affect the testability and if there are any cognitive benefits to one
approach over the other. 

% Theory
\chapter{Theory}\label{theory}

Software paradigms are ways to construct software. There exists many but the two
we will look at are functional programming and object-oriented programming.
Historically and today OOP has been the most popular language for the commercial
software industry. Java is an OOP language that is, in February 2019, the most
popular language in the software industry. Languages that are functional, like
Haskell and F\#, are less used.~\cite{tiobe} However functional concepts have
become more popular in multi-paradigm languages like Javascript.  For example
React, based on Functional Reactive Programming~\cite{reactisfrp}, has gained a
lot of popularity.~\cite{popularityreact} 

Exact definitions exist of OOP but not for Functional programming. Both
emphasise creating bug free programs. OOP emphasises encapsulating state into
\textit{objects} and message passing. Functional programs emphasise moving state
to the edges of the program, making the core logic of the program pure (defined
in Section~\ref{functionalprogramming}) and using immutable data.  Immutable
data is data whose state can not change once initialized. This is explained
further in Section~\ref{functionalprogramming} and Section~\ref{oop}.

While paradigms define how we build our applications, we still need design
patterns for structuring the source code. For instance, without design patterns
it is possible to couple dependencies with logic. This causes problems if we
later want to change the dependency since that means we also have to change the
logic. To increase maintainability, we want to use design patterns.  This
study will present the patterns for OOP and functional programming in their
respective section. When you create applications that you use once then
maintainability is not of high concern. However since the design patterns might
affect the metrics we will be looking at, we still employ them in this study.  

\section{Characteristics of Functional Programming}\label{functionalprogramming}

Functional programming is a software paradigm. While different definitions exist
of what it means, we define functional programming by the use of pure functions,
moving state to the outwards of the program, using trait-based polymorphism and
immutable data.

\subsubsection*{Purity}

When a function is pure it means that calling a function with the same arguments
will always return the same value and that it does not mutate any value. For
example if you have $f(x) = 2\cdot x$, then $f(2)$ will always return $4$. It
follows then that an impure functions is either dependant on some state or
mutates state in some way. For example, given $g(x) = currenttime \cdot x$,
$g(5)$ will yield a different value depending on what time it is called. This
makes it dependant on some state of the world. Or given $x=0$, $h()=x+1$. Then
$h()$ will yield $x=1$ and $(h \circ h)()$ will yield $x=2$, making it
impure.~\cite{wikipedia_pure}

\subsubsection*{Trait-based polymorphism}

In OOP we inherit classes that contain methods and
attributes.~\cite{Gamma:1995:DPE:186897} For functional programs, we instead
define classes that describe the actions that are possible. For example, a class
\texttt{Equality} could contain a function \texttt{isEqual} that checks if two
data types are equal. Then any data type that implements the interface Equality,
for example lists or binary trees, would be able to use the function isEqual.
This is known as type-classes in Haskell, mixins in Javascript or traits in
Scala.

\subsubsection*{Immutable data}

Immutable data is that that upon being initialized can not be changed. 

\subsubsection*{Moving state to the outwards of the program}

Even if functional programs emphasise purity certain applications still need to
deal with state somehow. For example a server would need to interact with a
database. Functional programs solve this by separating pure functions and
effectful functions. While various strategies exist, like Functional Reactive
Programming\footnote{Read more:
\url{en.wikipedia.org/wiki/Functional_reactive_programming}},
Dialogs\footnote{Read more:
\url{stackoverflow.com/questions/17002119/haskell-pre-monadic-i-o}} or
uniqueness types\footnote{Read more:
\url{https://en.wikipedia.org/wiki/Clean_(programming_language)}}, the one used
in Haskell, the language used in this thesis to construct the programs, is the
IO monad. For the uninitiated, one can think of Monads as a way to note which
functions are pure and which are effectful and managing the way they
intermingle. It also allows us to handle errors and state.\footnote{This is
simplified as Monads are notoriously difficult to explain.}. 

A strategy to further separate state and logic, one can construct a
three-layered architecture, called the three layer Haskell cake. Here, the
strategy is that one implements simple effectful functions, containing no logic
as a base layer. Then on a second layer one implements an interface that
implements a pure solution and one effectful solution. Then on the third layer
one implements the logic of the program in pure code. The way the second layer
is implemented is explained further in Section~\ref{interpreterpattern}. 

So while no exact definition of Functional programming exist, we here define it
as making functions pure and inheritance being based around functionality rather
than attributes.

\subsection{Interpreter pattern for testability}\label{interpreterpattern}

When structuring large functional applications there is no official way to do
it. For this study we will use a design pattern called the Interpreter pattern.
Informally, we can think of it as a way to create smaller composable compilers
that when added together make one big application. A compiler is a program that
takes some input, interprets the input and then does some output. A server, for
example, would take some request, interpret that request and then turn it into a
response. The server could integrate itself with the database, which would take
some query, interpret that query and then return an
object.~\cite{interpreterpattern}

To implement this pattern in Haskell we create an Abstract Syntax Tree (AST),
using a union type, of the program that contains all the available commands that
the program is capable of doing. See Figure~\ref{freeunion} for an example of a
to-do list AST. Once we have the AST we can encode the logic of the program as
instructions. Then the final step is to implement an interpreter for the program
that evaluates those commands. So if we have the commands in
Figure~\ref{freeunion}, we implement a function \texttt{eval} that takes a
command and computes some effectful code. The command \texttt{Add Item (Item ->
next)} could, for example, be executed as add an Item to a database.

\begin{figure}[H]
    \begin{lstlisting}
       data TodoList next
            = Add Item (Item -> next)
            | Mark Item next
            | Remove Item next
            | End --^ Terminates the program
    \end{lstlisting}
    \label{freeunion}
    \caption{AST for a to-do-list. We can derive a functor instances
    from ASTs for deriving Free instances.~\cite{commentarycompiler}}
\end{figure}


Hiding the implementation behind an AST allows us to separate effectful code
(like output a string or send a http request) with the logical instructions.
This simplifies our testing, since we can hide the environment (for example
database) behind an interface that we can swap out for testing. So we can
implement two interpreter functions, one for our real environment and one for
testing. 

\subsection{Extending applications with Free Monads}

\marginnote{{\footnotesize\textbf{Note:} \textit{This part assumes familiarity
with Haskell and uninitiated can ignore it.}}}

If we implement an interpreter for an AST we will run into the Expression
problem. The expression problem states:

\blockquote{The goal is to define a data type by cases, where one can add new
cases to the data type and new functions over the data type, without recompiling
existing code, and while retaining static type safety.}

This means, if we try to add another operation to our AST we would have to add
new cases to every function using that data type. For small programs this is not
a problem but as this scales it could mean adding cases to thousands of
functions.  To resolve this we will create a Free Monad for our AST. A Free
Monad is a Monad that we get ``for free'' if we instance an AST as a Functor.
With the do-notation a Free Monad gives an interface we can use to write our
program. So we start off by instancing it as a Functor (example in
Figure~\ref{freefunctor}). Functors for ASTs can automatically be derived using
the DeriveFunctor extension.~\cite{commentarycompiler}

\begin{figure}[H]
    \begin{lstlisting}
        instance Functor TodoList where
          fmap f (Add item k) = 
            Add item (f . k)
          fmap f (Mark item next) = 
            Set item (f next)
          fmap f End = 
            End
    \end{lstlisting}
    \label{freefunctor}
    \caption{A Functor instance for Figure~\ref{freeunion}. This can
    automatically be derived using DeriveFunctor.}
\end{figure}

From this we can define \texttt{type FreeAST = Free FunctorAst}. Lastly we lift
all the members of the union type as operations and create an interpreter. See
Figure~\ref{freeprogram} for a full example.

\begin{figure}[H]
    \begin{lstlisting}
        type TodoListF = Free TodoList

        add :: Item -> TodoListF Item
        add item = 
            liftF $ Add item id
        mark :: Item -> TodoListF ()
        mark item = 
            liftF $ Mark item ()
        end :: TodoListF r
        end = 
            liftF $ End

        eval :: TodoListF r -> IO r
        eval (Pure r) = return r
        eval (Free (Add item next)) = addItemToDb item >>= eval next
        eval (Free (Mark item next)) = markItem item >> eval next
        eval (Free End) = exitSuccess

        fakeItem :: Item

        exampleProgram :: TodoListF ()
        exampleProgram = 
            do  item <- add fakeItem 
                mark item 
                end

        main = 
            eval exampleProgram
    \end{lstlisting}
    \caption{An example implementation for our To-do list data type defined in
    Figure~\ref{freeunion}. Requires the haskell package ``free''
    (\protect\url{hackage.haskell.org/package/free-5.1})}
    \label{freeprogram}
\end{figure}

So to summarize, if we just implement an evaluate function for our AST we run
into the expression problem. To mitigate this we create Free monads from our
ASTs and to get an extendable interface that we can use with the do-notation. By
using the interpreter pattern, our program becomes independent from the
environment that it runs on. This allows us to easier test all parts of the
program.

\section{Object Oriented Programming}\label{oop}

Object-oriented programming (OOP) is a programming paradigm based around the
idea of objects that contain attributes and methods. The general idea is that
you model the world through objects, where the programming objects represent the
objects in the real world. A shopping cart might have an object ``Customer'' and
an object ``Shopping list'', for example. Inheritance is also an important
concept in OOP, where one object can inherent methods and attributes of another
object.  For example, given an object Animal with the attribute health and the
method attack, another object Alligator could inherit those
properties.~\cite{Gamma:1995:DPE:186897}

\subsection{SOLID principles}

If the design of a system is done poorly, it can lead to rotten design. Martin
Robert, a software engineer, claims that there are four big indicators of rotten
design.

\subsubsection*{Rigidity} Rigidity is the tendency for software to be difficult to
change. This makes it difficult to change non-critical parts of the software and
what can seem like a quick change takes a long time.

\subsubsection*{Fragility} Fragility is when the software tends to break when doing
simple changes. It makes the software difficult to maintain, with each fix
introducing new errors.

\subsubsection*{Immobility} Immobility is when it is impossible to reuse software from
other projects in the new project. So engineers discover that, even though they
need the same module that was in another project, too much work is required to
decouple and separate the desirable parts.

\subsubsection*{Viscosity} 

Viscosity comes in two forms: the viscosity of the environment and the viscosity
of the design. When the engineer makes changes they are often met with multiple
solutions. Some that preserve the design and some that are ``hacks''. So the
engineer can easily do the wrong thing. When the compile times are long the
engineers will attempt to make changes that do not cause long compile times.
This leads to viscosity in the environment.

To avoid creating rotten designs, Martin Robert proposes the SOLID guideline.
SOLID mnemonic for five design principles to make software more maintainable,
flexible and understandable. The SOLID guideline is

\begin{itemize}
    \item Single responsibility principle. Here, responsibility means ``reason
        to change''. Modules and classes should have one reason to change and no
        more.
    \item Open/Closed principle. States we should write our modules to be
        extended without modification of the original source code.
    \item Liskov substitution principle. Given a base class and an derived
        class derive, the user of a base class should be able to use the derived
        class and the program should function properly.
    \item Interface segregation principle. No client should be forced to depend
        on methods it does not use. The general idea is that you want to split
        big interfaces to smaller, specific ones.
    \item Dependency inversion principle. A strategy to avoid making our source
        code dependent on specific implementations is by using this principle.
        This allows us, if we depend on one third-party module, to swap that
        module for another one should we need to. This can be done by creating
        an abstract interface and then instance that interface with a class that
        calls the third-party operations.~\cite{martinrobert}
\end{itemize}

When we compare our solutions we want to test them after principles used in the
industry, as then we can use the results when making technical decisions for our
softwares. Using a SOLID architecture also helps us make programs that are not
as dependent on the environments, making them easier to test (as we can swap the
production environment to a test environment). When we investigate the
testability, it is important that we look at programs that are written in such a
way that all parts are easy to test. Thus choosing a SOLID architecture for our
OOP based programs will allow us to make more testable software.

%Method
\section{Measuring testability and complexity}\label{measuretestability}

With the definitions of the software paradigms, we can now look at how we can
evaluate them in regards of testability and complexity. Since we want to find
which of the two paradigms allows us to write the most maintainable software, it
is important that we have testable software so that finding defects is easier.
While we can not measure testability, by looking at the \textit{Cyclomatic
complexity} of our software we can find out how many tests would be needed to
test every possible outcome of the software. It follows then that if one of the
paradigms have a lower complexity, it would require less tests. If less tests
need to be written, it is logically sound to assume that the amount of lines of
code (LOC) for our software would be lower. Lower LOC correlates with less
defects in software.~\cite{defectloc} 

Looking at the testability of code for maintenance is not enough. While
testability is an important metric, it can also be important to look at other
factors like how cognitive complexity affects the maintenance. For instance, a
program written in the language Whitespace, an esoteric language consisting only
of whitespace\footnote{Description of Whitespace:
\url{https://en.wikipedia.org/wiki/Whitespace_(programming_language)}}, could
have a low LOC and cyclomatic complexity. But for humans knowing only the
English language, a text consisting of whitespace would be illegible. While
measuring cognitive complexity is hard, we can use a framework called Cognitive
dimensions to do a qualitative evaluation of software. We describe this further
in Section~\ref{cognitivedimensions}.


\subsection{Measuring testability: Cyclomatic Complexity}\label{cyclomaticcomplexity}

Cyclomatic complexity is a complexity measure that allows us to measure the
amount of paths through a program. The Cyclomatic complexity number is an upper
bound for the number of test cases required for full branch coverage of the
code. 

\theoremstyle{definition}
\begin{definition}
The cyclomatic number $v(G)$ of a graph G with $n$ vertices, $e$ edges and $p$
connected components is $v(G) = e - n + p$.
\end{definition}

\begin{theorem}
In a strongly connected graph $G$, the cyclomatic number is equal to the
maximum number of linearly independent circuits.~\cite{McCabe}
\end{theorem}

Informally, we can think of cyclomatic complexity as a way to measure the amount
tests a program needs to reach full branch coverage. We construct a graph that
branches out based on when the control flow in our source code branches out. For
example, given \texttt{f(bool) = if bool then ``hello''; else ``hola''} the
function \texttt{f} will either evaluate to ``hello'' or ``hola''. To get full
branch coverage for \texttt{f} we will therefore need two tests. The cyclomatic
complexity then becomes 2. If we construct a graph for the control flow of our
program, the nodes of the graph would represent processing tasks and edges
represent control flow between the nodes. 

\begin{figure}[H]
    \begin{lstlisting}
        void foo(void)
        {
          if (a)
            if (b) 
              x=1;
          else
              x=2;
         }
    \end{lstlisting}
    \caption{Multi if function foo.}\label{c1excode}
\end{figure}

If we have the code found in example figure~\ref{c1excode}. To calculate the
complexity of this function we first construct a graph as seen in
figure~\ref{fig:c1exgraph}. From the graph we find $n=4, e=5, p=2\Rightarrow
v(G)=e-n+p=5-4+2=3$ is the cyclomatic number.

\begin{figure}[H]
    \centering
    \begin{tikzpicture}
        \tikzset{vertex/.style = {shape=circle,draw,minimum size=2.5em}};
        \tikzset{edge/.style = {->,> = latex'}};
        \node[vertex] (a) at (2,8) {\texttt{if (a)}};
        \node[vertex] (b) at (0,6) {\texttt{if (b)}};
        \node[vertex] (c) at (4,6) {\texttt{x=2}};
        \node[vertex] (d) at (2,4) {\texttt{end}};
        \draw[edge] (a)  to (b);
        \draw[edge] (a)  to (c);
        \draw[edge] (b)  to (c);
        \draw[edge] (b)  to (d);
        \draw[edge] (c)  to (d);
    \end{tikzpicture}
    \caption{Cyclomatic complexity graph for Figure~\ref{c1excode}}\label{fig:c1exgraph}
\end{figure}

\subsection{Cyclomatic Complexity in Functional
Programming}

The definition of cyclomatic complexity in Section~\ref{cyclomaticcomplexity} is
not ideal for functional programming. Cyclomatic complexity is calculated by
creating graphs based on control flow operations such as while loops and if
statements. In functional programming everything is a function, thus the
cyclomatic complexity will always tend to 0 using this definition. So we define
a different method of calculating the cyclomatic complexity for functional
programs. 

\theoremstyle{definition} 
    \begin{definition} 
    The cyclomatic complexity number, in functional programming, is equal to
    1 plus the sum of the left hand side, called LHS, plus the sum of the
    right hand side, called RHS. RHS is the sum of the number of guards,
    logical operators, filters in a list comprehension and the pattern
    complexity in a list comprehension. LHS is equal to the pattern
    complexity.  The pattern complexity is equal to the number of
    identifiers in the pattern, minus the number of unique identifiers in
    the pattern plus the number of arguments that are not identifiers. In
    summary:

    \begin{lstlisting}
    Cyclomatic complexity = 1 + LHS + RHS

    LHS = Pattern complexity 

    Pattern complexity   
        = Pattern identifiers 
        - Unique pattern identifiers 
        + Number of arguments that are non identifiers

    RHS = Number of guards 
        + Number of Logical operators 
        + Number of filters in list comprehension 
        + Pattern complexity in list comprehension
    \end{lstlisting}
\end{definition}

Instead of cyclomatic graphs we instead construct flowgraphs, such as the one
seen in Figure~\ref{fig:cyclomaticfunctional} to model our function.

\begin{figure}[H]
    \begin{lstlisting}
    split :: (a -> Bool) -> [a] -> ([a], [a])
    split onCondition [] = ([], [])
    split onCondition (x:xs) =
        let 
            (ys, zs) = split onCondition xs
        in 
            if (onCondition x) then 
                (x:ys, zs)
            else 
                (ys, x:zs)
    \end{lstlisting}
    \label{split}
    \caption{Recursively split a list into two based on a given condition in
    Haskell. For example \texttt{split (>3) [1,2,3,4,5] =
    ([4,5],[1,2,3])}.}
\end{figure}

In Haskell $(x:xs)$ denotes an item $x$ at head of a list of items $xs$. Given
the Haskell code in Figure~\ref{split}. To calculate LHS we find two
pattern identifiers which are $onCondition$ and $(x:xs)$. there is one unique
pattern identifiers which is $(x:xs)$. There is also one non identifier
which is $[]$. We also find one guard, an if statement, and no
list comprehensions on RHS. Thus the cyclomatic complexity is $1+(2-1+1)+1=4$.
<-- NEEDS VERIFICATION FROM SOMEONE, THE SOURCE IS VERY VAGUE HERE\ldots

We do not count the $otherwise$ and $else$ clauses as a guard, just as how we
do not count the $else$ statement in normal procedural cyclomatic
complexity.~\cite{bergklaas}


\begin{figure}[H]
    \centering
    \begin{tikzpicture}
        \tikzset{vertex/.style = {shape=circle,fill=black,scale=0.5}};
        \draw[dashed] (0,0) -- (0,-5) (2,0) -- (2,-5) (4, 0) -- (4, -5) (6, 0)
        -- (6, -5);
        \node at (0,.3) {\small{Patterns}};
        \node at (2,.3) {\small{Guards}};
        \node at (4,.3) {\small{Expressions}};
        \node at (6,.3) {\small{Stop}};
        \node[vertex] (a) at (0,-0.3) {};
        \node[align=right, text width=3cm] at (-2,-0.3) {\texttt{[]}};
        \node[vertex] (b) at (4,-1.2) {};
        \draw[->, very thick] (a) to node [above] (TextNode) {T} (b) ;
        \node[vertex] (c) at (0,-1.5) {};
        \node[align=right, text width=3cm] at (-2,-1.5) {\texttt{(x:xs)}};
        \draw[->, very thick] (a) to node [left] (TextNode) {F} (c) ;
        \node[align=left, text width=3cm] at (8,-1.2) {\texttt{e1=([],[])}};
        \node[vertex] (d) at (2,-2.1) {};
        \draw[->, very thick] (c) to node [above] (TextNode) {T} (d) ;
        \node[align=right, text width=3cm] at (-2,-2.1) {\texttt{onCondition x}};
        \node[vertex] (e) at (4,-2.8) {};
        \draw[->, very thick] (d) to node [above] (TextNode) {T} (e) ;
        \node[align=left, text width=3cm] at (8,-2.8) {\texttt{e2=(x:yz, zs)}};
        \node[vertex] (f) at (4,-3.8) {};
        \draw[->, very thick] (d) to node [below] (TextNode) {F} (f) ;
        \node[align=left, text width=3cm] at (8,-3.8) {\texttt{e3=(yz, x:zs)}};
        \node[vertex] (g) at (6,-4.8) {};
        \draw[->, very thick] (e) to (g) ;
        \draw[->, very thick] (b) to (g) ;
        \draw[->, very thick] (f) to (g) ;
        \draw[very thick] (c) to (2, -4.0) to (g) ;
    \end{tikzpicture}
    \label{fig:cyclomaticfunctional}
    \caption{Flowgraph for split function, defined in Figure~\ref{split}.}
\end{figure}

Cyclomatic complexity allows us to look at programs in a quantitative way to
find how many tests we need to write for our programs to get full branch
coverage. A low complexity for a program means that less tests have to be
written.  Note that simply because a program has a low cyclomatic complexity it
does not imply that the program is easier to test. If a program is written in
such a way that it depends heavily on the environment it can also lead to
difficulty testing. We mitigate this difficulty by choosing a SOLID architecture
for OOP based programs and the interpreter pattern for functional based
programs.

\subsection{Mental complexity: Cognitive Dimensions}\label{cognitivedimensions}

Cognitive Dimensions is a framework for evaluating the usability of programming
languages and to find areas of improvements. It allows us to evaluate the
quality of a design and explore what future improvements could be made. As part
of the Cognitive Dimensions, 14 different Cognitive Dimensions of Notation
exist. A notation depends on the specific context, in this case the notation is
the languages themselves and their architecture. The author of the framework
recommends omitting the dimensions that are not applicable to the notation. We
give a brief description of the dimensions.

\subsubsection*{Viscosity}

How much work does it take to make small changes? How easy is the code to
refactor? If small changes requires consequent adjustments then that is a
problem. As a viscous system cause a lot more work for the user and break the
line of thought.

\subsubsection*{Visibility}

How easy is it to navigate the source code to find the parts that you want?

\subsubsection*{Hidden dependencies}

Are there hidden dependencies in the source code. Does a change in one part of
the source code lead to unexpected consequences in another part of the code.
Every dependency that matters to the user should be accessible in both
directions. 

\subsubsection*{Role-expressiveness}

How obvious is each sub-component of the source code to the solution as a whole?

\subsubsection*{Abstraction}

What are the levels of abstraction in the source code? Can the details be
encapsulated?

\subsubsection*{Secondary notation}

Are there any extra information being conveyed to the user in the source code?

\subsubsection*{Closeness of mapping}

By looking at the source code, how close do we find it to be to the case
we are solving?

\subsubsection*{Consistency}

Once Object-oriented procedural programming and Functional programming has been
learned. How much of the rest can the user guess successfully? 

\subsubsection*{Diffuseness or terseness}

How much space and symbols does the source code need to produce a certain result
or express a meaning?

\subsubsection*{Hard mental operations}

Where does the hard mental processing lie? Is it more when writing the source
code itself rather than solving the case, I.E. the semantic level? Does one
sometimes need to resort to pen and paper to keep track of what is happening?

\subsubsection*{Provisionality}

How easy is it to get feedback of something before you have completed the entire
system?

\subsubsection*{Progressive evaluation}

How obvious the role of each component of the source code in the solution as a
whole?

\subsubsection*{Error proneness}

To what extent does the software paradigm and language help minimise errors? Are
there any structures or complexities that lead to it being easier to make
errors?~\cite{GREEN1996131}\\

\noindent For this study we will investigate the following dimensions: 

\begin{itemize}
    \item Diffuseness or terseness
    \item Progressive evaluation
    \item Closeness of mapping
    \item Hard Mental Operations
    \item Visibility
    \item Hidden dependencies
    \item Abstraction
    \item Error-proneness 
\end{itemize}

\noindent We omit the other dimensions as related work concluded that the other
dimensions did not bring much weight when evaluating the different
paradigms.~\cite{euguenkiss}

In summary, cognitive dimensions allow us to look at different aspects of a
programming language to evaluate how complex they are cognitively. 

\chapter{Method: Case studies}

In the Chapter~\ref{theory} we defined what software paradigms are and how we
can evaluate their testability and their cognitive complexity using Cyclomatic
complexity and Cognitive dimensions. In Section~\ref{measuretestability} we also
explained how Cyclomatic complexity and Cognitive dimensions is tied to the
maintainability and testability of software. The aim of this study is to find if
the functional paradigm or the OOP paradigm is more maintainable than the other
and in which situations. This study will compare the cyclomatic complexity and
cognitive complexity by looking at three different cases. If we find that the
Cyclomatic complexity is lower in one of the paradigms or the other, we can
conclude that the amount of tests write for full branch coverage will be lower
for that paradigm than the other. By evaluating the paradigms using cognitive
dimensions we can find if one paradigm is easier to maintain for the developer.
The cases were chosen based on how they are often subproblems in bigger
applications. The solutions for functional programming will be implemented using
Haskell and for OOP we will use Java. 

Java is a programming language that is class-based and object-oriented. The aim
of the language is that you should be able to write the code and run it
anywhere.~\cite{java} Haskell is a purely functional programming language, it
also features lazy evaluation which allows you to compose functions easier. For
example, the function $three = take\; 3 \, \circ \, cycle\; 5$, where cycle is a
function that generates an infinitely long list consisting of a number, would
only compute the first three values, I.E. $[5,5,5]$, of the infinite list.  In a
non-lazy program it would never evaluate as $cycle$ would run
forever.~\cite{haskell} The reason these languages where chosen is because of
the authors familiarity with those languages.

\subsubsection*{Simplified chess game}

Chess is a famous game and in this report it is assumed that the reader know how
it works.\footnote{Rules of chess: \url{en.wikipedia.org/wiki/Rules_of_chess}}
Aim is to implement a simplified variant of it. This is not ordinary chess but a
simplified version:

\begin{itemize} 
    \item Only pawns and horses exist.
    \item You win by removing all the other players pieces.
\end{itemize}

The player should be able to do the following:

\begin{itemize} 
    \item List all available moves for a certain chess piece. 
    \item Move the chess piece to a given space
    \item Switch player after move
    \item Get an overview of the board
    \item Get an error when making invalid moves
\end{itemize}

To interact with the game, a user types commands through a command line prompt.
An example of basic interaction with the game is displayed in
Figure~\ref{chessexample}.

\begin{figure}[H]
    \centering
    \begin{lstlisting}
        > list (a,2)
        White pawn at (a,2)
        > move (a,2) (a,3)
        White pawn moved to (a,3)
        > listall
        White pawn at (a,3)
        White pawn at (b,2)
        ...
        White Horse at (a,1)
        ...
        Black pawn at (a,8)
        ...
    \end{lstlisting}
    \label{chessexample}
    \caption{An example of the interaction in the chess game.}
\end{figure}


\subsubsection*{To-do List}

A common task in programming is to create some kind of data store with
information. A to-do list is a minimal example of that. It consists of a list of
items that can be used to remember what to do later. The user should be able to:

\begin{itemize}
    \item Create a new item in the to-do list.
    \item Remove an item from the to-do list.
    \item Mark an item from the to-do list as done.
    \item See all items in the to-do list.
\end{itemize}

The interface to this program will be a text interface, displaying each item in
the todo list. To navigate and mark an item in the list the user presses up and
down and presses \texttt{x} to mark it.

\subsubsection*{Chat-bot engine}

Oftentimes when developing applications we have to deal with complex information
input. One of those cases is when we have chat-bots. Chat-bots are interactive
programs that respond with a text answer to the users input. For this
application we will implement the following:

\begin{itemize}
    \item Interpretor that can handle semi-complex inputs and deal with errors.
    \item Give answers to those inputs in form of text messages.
\end{itemize}

\noindent Once these programs have been constructed then cyclomatic complexity
can be evaluated by summing the cyclomatic complexity of each function. We can
also evaluate the cognitive complexity using the cognitive dimensions framework.
Thus we get can see if smaller subproblems in big applications require more
tests and have a bigger mental complexity depending on the different paradigms.

\chapter{Results}\label{results}
We worked so hard, yet achieved very little. TODO

\section{Limitations}\label{limitations}

TODO

\subsection{Improvements to implementation}
TODO

\section{future work}
\subsection{Relations to cardinality}
TODO

\bibliographystyle{ieeetr}
\bibliography{thesis}

\end{document}
